\Huge Conferences in an Era of Expensive Carbon

### 2019-12-26 main.tex
Nother round of polishing

\newif\ifdraft\drafttrue    % comments / todos for now
\newif\iflater\latertrue    % comments / todos for final version

@@

Indeed, scientific societies have an especially important role to
play, since, for many members, travel to conferences represents a
substantial or even dominant part of their individual contribution to
climate change.  A single round-trip flight from Philadelphia to Paris emits
the equivalent of about 1.8 tons of carbon dioxide (\COtwoE) per
passenger~\cite{carboncalculator}, which is a significant fraction of the
{\em total} yearly emissions for an average resident of the US (16.5 tons)
or Europe (7 tons)~\cite{emissions}.
Moreover, these emissions have no near-term technological
fix, as jet fuel is difficult to replace with renewable energy
sources~\cite{elec-air}.
% \iflater\bcp{Trimmed this sentence: Reducing the \COtwoE{} footprint
% of conferences will require a reduction in the footprint of the attendees'
% air travel.}\fi

How should ACM respond to these facts?  In 2016, ACM's Special
Interest Group on Programming Languages (SIGPLAN) convened an ad hoc Climate
Committee to consider this question~\cite{ClimateCommittee}.  After
investigating many options~\cite{ClimateCommitteReport}, we are putting
forward two concrete proposals. First, all ACM
conferences should publicly {\em account} for the {\COtwoE} emitted as a
result of putting them on---in particular, from travel to
the conference. Second, ACM should put a \emph{price} on
carbon in conference budgets, creating a steady pressure on organizers to reduce
their footprints.

@@

Our first proposal is modest: {\bf Every ACM-sponsored conference should
publicly report its carbon footprint}.
%
These reports should be collected in a central place, in a uniform format.
Most conferences' footprints will be dominated by participants' air travel,
but the data gathered should go beyond this to include ground
transportation, travel to in-person PC meetings, and estimated emissions
from hotel and food.

ACM should develop tools to gather and publicize this data.  For example,
SIGPLAN recently built an air-travel-focused {\COtwoE} calculator for
conferences\iflater~\cite{ClimateCommittee}.  Users can upload conference
registration data, and the calculator will estimate the {\COtwoE} cost of
air travel for those sufficiently far away. This calculator could be
expanded to account for other forms of conference emissions.

There is some the reputational risk to ACM in taking the step of publicizing
its carbon footprint: the numbers are likely to be high, and they may be
used to criticize both the ACM and the broader academic community.  But
making this information available is a crucial first step: we cannot manage
what we have not measured.  And ACM is well positioned to develop tools that
other academic societies can adopt.

@@

Similarly, SIGPLAN has been considering how to reduce emissions over the
past few years, informed by an accounting of its own {\COtwoE}
emissions. This discussion has led the organizers of two flagship
conferences (POPL and ICFP) to switch from in-person to online program
committee meetings, joining a trend among other SIGs, and has prompted
several conferences to increase investments in livestreaming and video
recording to support remote participation.

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{SIGPLAN-confs2.png}
\includegraphics[width=6in]{ParticipantsOriginAll2.png}
% \includegraphics[width=6in]{Distances.png}
\caption{Top: Carbon footprint per participant for travel to recent SIGPLAN
  conferences.  The smallest dot (ICFP
  14, in Gothenburg, Sweden) represents .9 tons of {\COtwoE} per participant; the
  largest (ICFP 16, in Nara, Japan) represents 1.94 tons per
  participant. Bottom: Breakdown of continent-of-origin for participants in
  each conference.  Colored bars represent percentages of participants whose
  home city is in each continent: blue for North America, orange for Europe,
  green for Asia.
% \iflater\bcp{Richard: I find this a
%     little hard to follow. Maybe you could simply label the different groups
%     of conferences by their location? I think the "(NA)", "(EU)", and
%     "(IAS)" are a bit hard to read and decode at first, so that I didn't
%     understand where each conference was located at first glance. (Adrianne
%     had the same response when I showed her.) Maybe just put text under
%     each, maybe with a horizontal bracket, that says "North America,"
%     "Europe," and "Rest of world," with the bracket and/or words color-coded
%     as well.}\fi
%
% Bottom: Distances traveled by participants, one way, in Km; middle dashed
% line is the median travel distance.
}
\label{fig:confs}
\end{figure}

@@

However, while public accounting of {\COtwoE} will encourage easy
reductions, it is not likely, by itself, to induce major shifts in behavior.
%
Science is a fundamentally social process, and the conference system
accelerates scientific research through high-bandwidth interaction,
direct dissemination of results, network building, and serendipitous
cross-fertilization.
%
Organizers and attendees will naturally be reluctant to consider changes
that might threaten these benefits.

@@

To illustrate the challenges, consider the problem of choosing a conference
location that minimizes emissions from participants' travel.  Using recent
registration data from four \mbox{SIGPLAN} conferences, Figure~\ref{fig:confs}
shows two ways of looking at the relation between locations and emissions.
%
The top diagram shows an
estimated per-participant {\COtwoE} footprint for each instance of each
conference over the past ten years (excluding a few for which we
had trouble getting data), with larger dots representing higher emissions.
Eyeballing this diagram, it seems that carbon-conscious
organizers should hold all of these conferences in either the northeast US or
western Europe every year.  But the bottom visualization tells a
different story.  The horizontal colored bar at the bottom represents
the continent on which each conference was held, and each vertical bar gives
a breakdown of the participants in that conference, colored according to the
continent of their work address.  From a superficial glance
at the colors, it is clear that, though a minority come from far away, the
majority of participants in each conference are local to the region where the
conference is being held.
% \iflater\bcp{That point could be strengthened by adding some
%   numbers (just in the text, not a figure) about what percentage of
%   conference attendees attend even one additional instance of the same
%   conference...}\fi{}
 This suggests that always locating conferences in the
same one or two places would significantly impact the diversity of the
research community by discouraging participation from other parts of the
world.  Indeed, one might conclude that, from the point of view of
strengthening the research community, conferences should move around as
much as possible!
%
%
% \bcp{Explain the two diagrams.  Discuss how they can be used to draw
%   contradictory conclusions.  E.g., one of the many footprint-reduction
%   measures that have been discussed in SIGPLAN~\cite{ClimateCommitteReport}
%   is locating conferences near their historical ``center of mass.''  Looking
%   at the top diagram, one might thus be tempted to locate all conferences in
%   the northeast US or western Europe.  But the bottom figure shows that much
%   of the participation in each instance of a conference is local; always
%   holding a conference at its ``center of mass'' will damage local
%   communities everywhere else.  (Maybe this would be a good place to refer
%   to~\cite{ClimateCommitteReport}---we can enumerate some of the
%   alternatives that one might have hoped the data would help us choose
%   between.)}
% %
% \bcp{Maybe we can replace the bottom part of the figure by some clearer way
%   of visualizing the observation that much of the attendance at each
%   conference is local?  E.g., what if, instead of color-coding the different
%   sections of the bars by continent, we instead color-coded them by distance
%   traveled (in buckets of 500 miles, say)?}
%
These disparate perspectives suggest that reducing conference emissions may
require genuinely painful compromises.  The impulse to ignore the issue is
entirely understandable.

@@

\SECTION{Put a Price on Carbon}
%
This dilemma is a microcosm of one faced by all of society. To address it,
many policy experts advocate using some form of carbon pricing to impose a
concrete, immediate cost on
emissions~\cite{carbonprice}. Doing so makes manifest the hidden
environmental cost of emissions, incentivizing
{\COtwoE}-reducing changes without mandating exactly which ones, and thus
allowing for creative and efficient responses. Continuing the junk-food
analogy,
some municipalities including New York City and Philadelphia have imposed a
per-calorie tax on soft drinks, with some studies finding that doing so
indeed reduced consumption~\cite{sodatax}.

Thus, our second proposal is that {\bf ACM should impose a surcharge on
  conferences based on their carbon footprint.}  The charge should start low
and increase steadily and predictably, year on year.
%
Conference organizers can then choose how best to balance their
budgets---whether
by decreasing per-participant emissions, decreasing (physical)
participation, increasing registration fees, soliciting additional corporate
sponsorship, or other means.
In this process, a primary concern
should be to find ways of {reducing} the financial burden of such a
surcharge on those who may be disproportionately affected by it---students
without grant support, participants from developing areas, etc.
Wealthier participants should subsidize the carbon footprints of less
wealthy ones.

Ideally, at some point, governments will impose carbon pricing uniformly,
and all {\COtwoE}-intensive activities will have to pay it. But ACM can send a
strong message about the importance of this issue---and get ahead of the
coming changes---by acting now.

@@

\PARAGRAPH{Precipitating change}
%
What should ACM do with the funds collected from this surcharge?  One
obvious possibility is purchasing \emph{carbon offsets}~\cite{CarbonOFfsetReport}.
%
A carbon offset is sold by a vendor, who uses the funds to finance an
activity that permanently removes or avoids emitting some amount of
greenhouse gases.  The veracity and permanence of this activity is certified
by a watchdog organization. (E.g., planting trees is typically not a certifiable
activity, since it's hard to guarantee they won't be cut down, but
installing methane capture devices on landfill sites or buying
fuel-efficient stoves to replace open-fire cooking in poor communities can
be.)
Many organizations, including companies such as Google, Dell, Microsoft,
General Motors, Delta Airlines, Lyft, and Expedia, as well as universities,
academic societies, and even energy companies such as Exxon, now use
carbon offsets to reduce their
net footprint. ACM conferences should consider doing the same, and the
purchases should be
included in the public accounting we are proposing.  Indeed, ACM recently
declared that conferences can offer attendees the option to pay for an
offset during registration.  Other scientific societies and conferences are
going further, rolling a carbon offset into the registration fee for every
participant.

Beyond buying offsets, one can
imagine many good uses for the funds generated from a carbon surcharge: defraying the costs of
virtualizing conferences (livestreaming, etc.), and supporting ``green''
computing research. As an example of the last, ACM could help fund a
cross-cutting research initiative specifically aimed at understanding how to
best replace or approximate the socializing and networking aspects of
conferences in a virtual setting.

Ultimately, however, carbon offsets and other ``good works'' cannot
substitute for real reductions in emissions~\cite{anderson2012inconvenient,
  carbon-offsets-are-not-our-get-out-jail-free-card}: they are, at best, a
short-term expedient that buys a little time to develop strategies for more
difficult cuts.
%
Indeed, the main goal of carbon pricing should be to stimulate creative
rethinking of the conference model itself---for example, seriously
considering alternatives such as rapid-turnaround journal-only publishing
models, once-a-year megaconferences, and entirely virtual
conferences~\cite{NCN}.
%
A potential sticking point is that some of these will significantly reduce
conference revenues, in turn impacting the income stream of ACM itself; this
could make emissions reduction politically problematic unless ACM's
conference-focused business model is also adjusted.

@@

Another key issue in implementation of this second proposal will be how to
set a price that reflects the true social cost of carbon, without unduly
harming the scientific community.
%
Initial data gathering will play a role, and a steady and predictable annual
ramp on the price is a necessary component, but setting both the initial
price and the slope of the ramp will likely be difficult political
decisions.
%
One measurable target, in line with the latest evidence from climate
science, would be to tune the parameters with a goal of halving actual
emissions every decade from now on, following a recently proposed ``Carbon
Law''~\cite{Rockstrom1269}.

@@

The climate crisis is too urgent to leave to world leaders to address at
their own pace: Organizations at every scale, including ACM, must confront
their own contributions, raise awareness among their members, and establish
new ways of doing business in the lower-carbon future that is now upon us.
We in ACM should do our part by mandating public accounting of conference
carbon footprints and by putting a concrete price on the carbon that we use.

@@

@@

\iflater
\newpage
\section*{Permissions for figures}

>>>>

\newif\ifdraft\drafttrue    % comments / todos for now
\newif\iflater\laterfalse    % comments / todos for final version

@@

Indeed, scientific societies have an especially important role to
play, since, for many members, travel to conferences represents a
substantial or even dominant part of their individual contribution to
climate change.  A single round-trip flight from Philadelphia to Paris emits
the equivalent of about 1.8 tons of carbon dioxide (\COtwoE, or informally
``carbon'') per
passenger~\cite{carboncalculator}. This is a significant fraction of the
{\em total} yearly emissions for an average resident of the US (16.5 tons)
or Europe (7 tons)~\cite{emissions}.
Moreover, these emissions have no near-term technological
fix, as jet fuel is difficult to replace with renewable energy
sources~\cite{elec-air}.

How should ACM respond to these facts?

In 2016, ACM's Special
Interest Group on Programming Languages (SIGPLAN) convened an ad hoc Climate
Committee to consider this question~\cite{ClimateCommittee}.  After
investigating many options~\cite{ClimateCommitteReport}, we are putting
forward two concrete proposals. First, all ACM
conferences should publicly {\em account} for the {\COtwoE} emitted as a
result of putting them on---in particular, from travel to
the conference. Second, ACM should put a \emph{price} on
carbon in conference budgets, creating a steady pressure on organizers to reduce
their footprints.

@@

Our first proposal is modest: {\bf Every ACM-sponsored conference should
publicly report its carbon footprint}.
%
These reports should be collected in a central place, in a uniform format.
Most conferences' footprints will be dominated by participants' air travel,
but the data gathered should go beyond this to include ground
transportation, travel to in-person program committee meetings, and
estimated emissions from hotels and food.

ACM should develop tools to gather and publicize this data.  For example,
SIGPLAN recently built an air-travel-focused carbon calculator for
conferences~\cite{ClimateCommittee}.  Users can upload conference
registration data, and the calculator will estimate the {\COtwoE} cost of
air travel for those sufficiently far away.
% This calculator could be
% expanded to account for other forms of conference emissions.

There is some the reputational risk to ACM in taking the step of publicizing
its carbon footprint: the numbers are likely to be high, and they may be
used to criticize both the ACM and the broader academic community.  But
making this information available is a crucial first step: we cannot manage
what we have not measured.

@@

Similarly, SIGPLAN has been considering how to reduce emissions, informed by
an accounting of its own carbon footprint. This discussion has led the
organizers of two flagship
conferences (POPL and ICFP) to switch from in-person to online program
committee meetings, joining a trend among other SIGs, and has prompted
several conferences to increase investments in livestreaming and video
recording to support remote participation.

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{SIGPLAN-confs2.png}
\includegraphics[width=6in]{ParticipantsOriginAll2.png}
% \includegraphics[width=6in]{Distances.png}
\caption{Top: Carbon footprint per participant for travel to recent SIGPLAN
  conferences.  The smallest dot (ICFP
  14, in Gothenburg, Sweden) represents .9 tons of {\COtwoE} per participant; the
  largest (ICFP 16, in Nara, Japan) represents 1.94 tons per
  participant. Bottom: Breakdown of continent-of-origin for participants in
  each conference.  Colored bars represent percentages of participants whose
  home city is in each continent: blue for North America, orange for Europe,
  green for Asia.
% \bcp{Richard: I find this a
%     little hard to follow. Maybe you could simply label the different groups
%     of conferences by their location? I think the "(NA)", "(EU)", and
%     "(IAS)" are a bit hard to read and decode at first, so that I didn't
%     understand where each conference was located at first glance. (Adrianne
%     had the same response when I showed her.) Maybe just put text under
%     each, maybe with a horizontal bracket, that says "North America,"
%     "Europe," and "Rest of world," with the bracket and/or words color-coded
%     as well.}\fi
%
% Bottom: Distances traveled by participants, one way, in Km; middle dashed
% line is the median travel distance.
}
\label{fig:confs}
\end{figure}

@@

However, while public accounting of emissions will encourage easy
reductions, it is not likely, by itself, to induce major shifts in behavior.
%
Science is a fundamentally social process, and the conference system
accelerates scientific research through high-bandwidth interaction,
direct dissemination of results, network building, and serendipitous
cross-fertilization.
%
Organizers and attendees will naturally be reluctant to consider changes
that might threaten these benefits.

@@

To illustrate the challenges, consider the problem of choosing a conference
location that minimizes emissions from participants' travel.  Using recent
registration data from four \mbox{SIGPLAN} conferences, Figure~\ref{fig:confs}
shows two ways of looking at the relation between locations and emissions.
%
The top diagram shows an
estimated per-participant {\COtwoE} footprint for each instance of each
conference over the past ten years (excluding a few for which we
had trouble getting data), with larger dots representing higher emissions.
Eyeballing this diagram, it might seem that carbon-conscious
organizers should hold all of these conferences in either the northeast US or
western Europe every year.  But the bottom visualization tells a
different story.  The horizontal colored bar at the bottom represents
the continent on which each conference was held, and each vertical bar gives
a breakdown of the participants in that conference, colored according to the
continent of their work address.  A glance
at the colors makes clear that, though a minority come from far away, the
majority of participants in each conference are local to the region where the
conference is being held.
%
This suggests that always locating conferences in the
same one or two places would significantly impact the diversity of the
research community by discouraging participation from other parts of the
world.  Indeed, one might conclude that, from the point of view of
strengthening the research community, conferences should move around as
much as possible!
%
%
% \bcp{Explain the two diagrams.  Discuss how they can be used to draw
%   contradictory conclusions.  E.g., one of the many footprint-reduction
%   measures that have been discussed in SIGPLAN~\cite{ClimateCommitteReport}
%   is locating conferences near their historical ``center of mass.''  Looking
%   at the top diagram, one might thus be tempted to locate all conferences in
%   the northeast US or western Europe.  But the bottom figure shows that much
%   of the participation in each instance of a conference is local; always
%   holding a conference at its ``center of mass'' will damage local
%   communities everywhere else.  (Maybe this would be a good place to refer
%   to~\cite{ClimateCommitteReport}---we can enumerate some of the
%   alternatives that one might have hoped the data would help us choose
%   between.)}
% %
% \bcp{Maybe we can replace the bottom part of the figure by some clearer way
%   of visualizing the observation that much of the attendance at each
%   conference is local?  E.g., what if, instead of color-coding the different
%   sections of the bars by continent, we instead color-coded them by distance
%   traveled (in buckets of 500 miles, say)?}
%
These disparate perspectives suggest that significantly reducing conference
emissions may
require genuinely painful compromises.  The impulse to ignore the issue is
entirely understandable.

@@

\SECTION{Put a Price on Carbon}
%
This dilemma is a microcosm  of one faced by all of society. To address it,
many policy experts advocate using some form of carbon pricing to impose a
concrete, immediate cost on
emissions~\cite{carbonprice}. Doing so makes manifest the hidden
environmental cost of emissions, incentivizing
{\COtwoE}-reducing changes without mandating exactly which ones, and thus
allowing for creative and efficient responses. Continuing the junk-food
analogy,
some municipalities including Berkeley and Philadelphia have imposed a
per-calorie tax on soft drinks; studies found that doing so
significantly reduced consumption~\cite{sodatax}.

Thus, our second proposal is that {\bf ACM should impose a surcharge on
  conferences based on their carbon footprint.}  The charge should start low
and increase steadily and predictably, year on year.
%
Conference organizers can then choose how best to balance their
budgets---whether
by decreasing per-participant emissions, decreasing (physical)
participation, increasing registration fees, soliciting corporate
sponsorship, or other means.
In this process, a primary concern
should be to find ways of {reducing} the financial burden of such a
surcharge on those disproportionately affected by it---students
without grant support, participants from developing areas, etc.
Well-funded participants should subsidize the carbon surcharges of less
wealthy ones.

Ideally, at some point, governments will impose carbon pricing uniformly,
and all carbon-intensive activities will have to pay it. But ACM can send a
strong message about the importance of this issue---and get ahead of the
coming changes---by acting now.

@@

\PARAGRAPH{Precipitating change}
%
What should ACM do with the funds collected from this surcharge?  One
obvious possibility is purchasing \emph{carbon offsets}~\cite{CarbonOFfsetReport}.
%
A carbon offset is sold by a vendor, who uses the funds to finance an
activity that permanently removes or avoids emitting some amount of
greenhouse gases.  The veracity and permanence of this activity is certified
by a watchdog organization. (E.g., planting trees is often considered not to
be a certifiable
activity, since it's hard to guarantee they won't be cut down;
reductions from
installing methane capture devices on landfill sites or buying
fuel-efficient stoves to replace open-fire cooking in poor communities are
easier to predict.)
Many organizations, including companies such as Google, Dell, Microsoft,
General Motors, Delta Airlines, Lyft, and Expedia, as well as universities,
academic societies, and even energy companies such as Exxon, now use
carbon offsets to reduce their
net footprint. ACM conferences should consider doing the same, and the
purchases should be
included in the public accounting we are proposing.
% For example\bcp{beter
%   transition?}, ACM recently
% declared that conferences can offer attendees the option to pay for an
% offset during registration.  Other scientific societies and conferences are
% going further, rolling a carbon offset into the registration fee for every
% participant.

Beyond buying offsets, one can
imagine many good uses for the funds generated from a carbon surcharge: defraying the costs of
virtualizing conferences (livestreaming, etc.), and supporting ``green''
computing research~\cite{rolnick2019tackling}.  As an example
of the last, ACM could help fund a
cross-cutting research initiative specifically aimed at understanding how to
best replace or approximate the socializing and networking aspects of
conferences in a virtual setting.

Ultimately, however, carbon offsets and other ``good works'' cannot
substitute for real reductions in emissions~\cite{anderson2012inconvenient,
  carbon-offsets-are-not-our-get-out-jail-free-card}: they are, at best, a
short-term expedient that buys time to agree on more
difficult cuts.
%
Indeed, the main goal of carbon pricing should be to stimulate creative
rethinking of the conference model itself---for example, seriously
considering alternatives such as rapid-turnaround journal-only publishing
models, yearly megaconferences, and entirely virtual
conferences~\cite{NCN}.
%
A potential sticking point is that some of these will significantly reduce
conference revenues, in turn impacting the income stream of ACM itself; this
could make emissions reduction politically problematic unless ACM's
conference-focused business model is also adjusted.

@@

Another key issue in implementing this second proposal will be how to
set a price that reflects the true social cost of carbon, without unduly
harming the scientific community.
%
Initial data gathering will play a role, and a steady and predictable annual
increase is a necessary component, but setting both the initial
price and the slope of the ramp will likely be difficult political
decisions.
%
One measurable target, in line with the latest evidence from climate
science, would be to tune the parameters with a goal of halving actual
emissions every decade from now on, following a recently proposed ``Carbon
Law''~\cite{Rockstrom1269}.

@@

The climate crisis is too urgent to leave to world leaders to address at
their own pace: Organizations at every scale, including ACM, must confront
their own contributions, raise awareness and foster discussion among their
members~\cite{acm-climate}, and establish new ways of doing business in the
lower-carbon future that is now upon us.  We in ACM should do our part by
mandating public accounting of conference carbon footprints and by putting a
concrete price on the carbon that we use.

@@

\iflater

@@

\newpage
\section*{Permissions for figures}
